Rapid and Accurate Global PTM Discovery (G-PTM-D) Using Post-Acquisition Spectral Calibration and Multi-Notch Searches

### 2017-07-10 paper.tex
minor
dmkfdsjf
yeah

Correct identification of protein post-translational modifications (PTMs) is crucial to understanding many aspects of protein function in biological processes.
G-PTM-D\cite{Li_2016} is a recently developed tool for global identification and localization of PTMs.
Spectral file calibration prior to applying G-PTM-D, and algorithmic enhancements in the peptide database search shown here significantly increase the accuracy and speed of PTM identification.
We enhance G-PTM-D by using targeted multi-notch searches, and demonstrate its effectiveness in identification of numerous types of PTMs, including high-mass modifications such as glycosylations.
The changes described in this work lead to a 20\% increase in the number of identified modifications and an order of magnitude decrease in the search time.
Targeted multi-notch searches are also shown to be useful in contexts other than G-PTM-D, producing superior results when used instead of standard narrow-mass and wide-mass searches.
\end{abstract}

@@

Many types of post-translational modifications (PTMs) are known, and databases containing detailed information about such modifications are readily available\cite{Creasy_2004}.
Information about a modification often includes the chemical or isotopic composition, mass, specificity to certain residues, and possible restriction of placement to peptide or protein termini.
Protein databases such as UniProt include information on the localization of specific modifications to specific residues in the protein, but this information is incomplete.
Despite the extensive cataloging of the different possible protein modifications, their comprehensive discovery and identification in complex biological samples has continued to pose a challenge for proteomics\cite{Olsen_2013}.

Numerous procedures for identification and localization of PTMs from "bottom-up" tandem mass spectrometric datasets exist, and those address primarily common modifications (e.g. phosphorylations) in well-studied systems.
Global discovery tools, such as MODa\cite{Na_2011} are also available.
G-PTM-D\cite{Li_2016} is one recently described bioinformatics tool for high confidence identification and localization of a wide variety of different PTMs.
The G-PTM-D workflow consists of three stages: 1) A wide-mass database search\cite{Chick_2015, Na_2011}, that provides mass differences between the identified peptides and the observed precursor peptide masses (hypothesized to differ in mass due to the presence of a PTM);
2) For those peptides for which the mass difference corresponds to the mass of a known PTM, a database augmentation step adds plausible localized PTMs to the corresponding protein entries in the search database;
3) A final standard narrow-mass search of the augmented database to identify both modified and unmodified peptides subject to the standard FDR threshold (e.g. 1\% FDR).

@@

\begin{figure}[H]
  \includegraphics[scale=0.5]{diagram.eps}
  \caption{Workflows}
  \label{fgr:diagram}
\end{figure}

@@

We developed a modified version of the Morpheus software for bottom-up spectral database searching\cite{Wenger_2013} that we call MetaMorpheus\footnote{Free and Open Source software available at \url{https://github.com/smith-chem-wisc/MetaMorpheus}}, which integrates the database search procedure with spectral calibration and the enhanced G-PTM-D workflow.
Three multi-fraction mammalian datasets (Jurkat cell, C57BL/6J (B6) and CAST/EiJ (CAST) mouse islet lysates) with deep proteome coverage were used to evaluate performance.
These tryptically digested cell lysates correspond to 28 fractions from human Jurkat cells, and 18 fractions from Mouse islets, and are described in detail in~\cite{Shortreed_2015, Cesnik_2016}.

@@

A critical parameter for peptide identification is mass accuracy\cite{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories: External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\cite{Olsen_2005}); and spectral calibration subsequent to the MS experiment (post-acquisition spectral calibration).
We developed a post-acquisition calibration procedure that builds upon the software lock mass concept\cite{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.
The spectral calibration procedure consists of two steps: Step 1: A narrow tolerance (e.g. 10 ppm precursor, 0.01 Dalton fragment) database search is performed on the dataset.
This yields a set of peptide identifications subject to a desired false discovery rate (e.g 1\%; based upon the target:decoy strategy\cite{Elias_2007}).
Step 2: The identifications from Step 1 are used to extract peak matches from the spectra, and a peak shifting procedure is performed.
The calibration algorithm is described in detail in Supplementary Material 1.

@@

\begin{figure}
  \includegraphics{fig2-calibrationQuality.eps}
  \caption{Histograms demonstrating the effect of calibration on PSM mass errors}
  \label{fgr:figure1}
\end{figure}

@@

\begin{figure}
\includegraphics{fig1-searchTypes.eps}
\caption{Search Modes}
\label{fig:fig1-searchtypes}
\end{figure}

@@

\begin{figure}
\caption{Mouse histograms around 0}
\label{fig:fig3mouse-1012}
\includegraphics{fig3mouse-1012}
\end{figure}

\begin{figure}
\caption{Jurkat histograms around 0}
\label{fig:fig3jurkat-1012}
\includegraphics{fig3jurkat-1012}
\end{figure}

@@

Chick~\citep{Chick_2015} showed that by significantly relaxing the parent mass tolerance in bottom-up searches, an incredible variety of unforeseen peptides were being fragmented and could be identified.
In particular, PTM-containing peptides could be identified from those MS2 fragments that did not possess the modification and the type of PTM could be inferred from differences in precursor and parent mass.
A major limitation of the Chick work was the extremely long search times because of the vast number of theoretical candidate peptides that were evaluated for each and every MS2 spectrum.
Recently, Kong\cite{Kong_2017} reported a new index-based search strategy (MSFragger) that radically improved search speeds for wide parent mass tolerance searches.
Search times for +/- 500 Da parent tolerances went from six years to six seconds.
The output of both search strategies is a list of PSMs, where the best matching unmodified theoretical peptide was found for each MS2 fragmentation spectrum and the mass difference between the theoretical and experimental precursor masses was determined.
The mass differences can be clustered into histogram peaks\citep{Rodriguez_2014}, but interpretation of the mass difference was left to the user.
PTMs corresponding to the reported mass difference can frequently be inferred; such as when the mass difference reported matches a known PTM mass (e.g. 79.966331 Da for phosphorylation).
However, mass differences can arise from not only one modification, but frequently from two or more modifications, some of which increase the mass and the peptide and some of which decrease it. In addition, small missed cleavages are sometimes reported mistakenly as mass differences. Also, some mass differences, such as those matching to amino acid masses, are easily interpreted but suffer from high local FDR and should only be used with caution. Finally, seemingly uninterpretable mass differences can be reported when the best matching peptide has an actual precursor mass different from the precursor mass selected by the mass spectrometer. The range of masses selected for fragmentation in a tandem mass spectrometer is commonly +/- 2 Da. Therefore, multiple peptides can be introduced to the cell and co-fragmented. This often results in peptide identification of a less abundant peptide with mass slightly larger or smaller than the selected precursor.
The unconstrained utilization of mass difference and acceptance of the theoretical best matching sequence as true can result in a high number of false positive assignments.

@@

\begin{figure}	
	\centering
	\begin{subfigure}[t]{3.1in}
		\centering
		\includegraphics[scale=0.5]{figPhosphorylationPre.eps}
		\label{fig:1a}		
	\end{subfigure}
	\quad
	\begin{subfigure}[t]{3.1in}
		\centering
		\includegraphics[scale=0.5]{figPhosphorylationCalib.eps}
		\label{fig:1b}
	\end{subfigure}
	\caption{Histograms of mass differences observed around 80 Daltons when using calibrated and uncalibrated spectra}\label{fig:phosphoHistograms}
\end{figure}

@@

Since the XML database is annotated with PTMs from the UniProt PTM list, it is natural to use it in the augmentation step as well.
It has a few significant drawbacks: Some mass values do not correspond to actual mass shifts, lability of modifications is not taken into account, and many PTMs are missing, including adducts and glycosylations.
Notch searches allow using expanded modification lists that include chemical modifications such as adducts and large mass PTMs such as glycosylations.
Even though adducts arising from experimental artifacts are not ultimately interesting to biologists, not including them as an option decreases the total number of identified proteins and PTMs.
A curated set of notches and the corresponding modifications is given in Appendix A.
The overall identification rate increased from 164,697 to 179,345 peptide-spectral matches (PSMs), which in turn increased the number of modified peptides by an additional 20\%.
We identified hundreds of glycosylated peptides in these unenriched samples, with many of these modifications exceeding 1000 Da.

>>>>

Correct identification of protein post-translational modifications (PTMs) is crucial to understanding many aspects of protein function in biological processes.
G-PTM-D\citep{Li_2016} is a recently developed tool for global identification and localization of PTMs.
Spectral file calibration prior to applying G-PTM-D, and algorithmic enhancements in the peptide database search shown here significantly increase the accuracy and speed of PTM identification.
We enhance G-PTM-D by using targeted multi-notch searches, and demonstrate its effectiveness in identification of numerous types of PTMs, including high-mass modifications such as glycosylations.
The changes described in this work lead to a 20\% increase in the number of identified modifications and an order of magnitude decrease in the search time.
Targeted multi-notch searches are also shown to be useful in contexts other than G-PTM-D, producing superior results when used instead of standard narrow-mass and wide-mass searches.
\end{abstract}

@@

Many types of post-translational modifications (PTMs) are known, and databases containing detailed information about such modifications are readily available\citep{Creasy_2004}.
Information about a modification often includes the chemical or isotopic composition, mass, specificity to certain residues, and possible restriction of placement to peptide or protein termini.
Protein databases such as UniProt include information on the localization of specific modifications to specific residues in the protein, but this information is incomplete.
Despite the extensive cataloging of the different possible protein modifications, their comprehensive discovery and identification in complex biological samples has continued to pose a challenge for proteomics\citep{Olsen_2013}.

Numerous procedures for identification and localization of PTMs from "bottom-up" tandem mass spectrometric datasets exist, and those address primarily common modifications (e.g. phosphorylations) in well-studied systems.
Global discovery tools, such as MODa\citep{Na_2011} are also available.
G-PTM-D\citep{Li_2016} is one recently described bioinformatics tool for high confidence identification and localization of a wide variety of different PTMs.
The G-PTM-D workflow consists of three stages: 1) A wide-mass database search\citep{Chick_2015, Na_2011}, that provides mass differences between the identified peptides and the observed precursor peptide masses (hypothesized to differ in mass due to the presence of a PTM);
2) For those peptides for which the mass difference corresponds to the mass of a known PTM, a database augmentation step adds plausible localized PTMs to the corresponding protein entries in the search database;
3) A final standard narrow-mass search of the augmented database to identify both modified and unmodified peptides subject to the standard FDR threshold (e.g. 1\% FDR).

@@

\begin{figure}[H]
  \includegraphics{paperDiagram.png}
  \caption{The three steps of the G-PTM-D workflow are indicated in the light beige color.}
  \label{fgr:diagram}
\end{figure}

@@

We developed a modified version of the Morpheus software for bottom-up spectral database searching\citep{Wenger_2013} that we call MetaMorpheus\footnote{Free and Open Source software available at \url{https://github.com/smith-chem-wisc/MetaMorpheus}}, which integrates the database search procedure with spectral calibration and the enhanced G-PTM-D workflow.
Three multi-fraction mammalian datasets (Jurkat cell, C57BL/6J (B6) and CAST/EiJ (CAST) mouse islet lysates) with deep proteome coverage were used to evaluate performance.
These tryptically digested cell lysates correspond to 28 fractions from human Jurkat cells, and 18 fractions from Mouse islets, and are described in detail in~\citep{Shortreed_2015, Cesnik_2016}.

@@

A critical parameter for peptide identification is mass accuracy\citep{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories: External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\citep{Olsen_2005}); and spectral calibration subsequent to the MS experiment (post-acquisition spectral calibration).
We developed a post-acquisition calibration procedure that builds upon the software lock mass concept\citep{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.
The spectral calibration procedure consists of two steps: Step 1: A narrow tolerance (e.g. 10 ppm precursor, 0.01 Dalton fragment) database search is performed on the dataset.
This yields a set of peptide identifications subject to a desired false discovery rate (e.g 1\%; based upon the target:decoy strategy\citep{Elias_2007}).
Step 2: The identifications from Step 1 are used to extract peak matches from the spectra, and a peak shifting procedure is performed.
The calibration algorithm is described in detail in Supplementary Material 1.

@@

%\begin{figure}
%  \includegraphics{calib1.png}
 % \caption{Histograms demonstrating the effect of calibration on PSM mass errors}
 % \label{fgr:figure1}
%\end{figure}

@@

%\begin{figure}
%\includegraphics{fig1-searchTypes.png}
%\caption{Search Modes}
%\label{fig:fig1-searchtypes}
%\end{figure}

@@

%\begin{figure}
%\caption{Mouse histograms around 0}
%\label{fig:fig3mouse-1012}
%\includegraphics{fig3mouse-1012.png}
%\end{figure}

%\begin{figure}
%\caption{Jurkat histograms around 0}
%\label{fig:fig3jurkat-1012}
%\includegraphics{fig3jurkat-1012.eps}
%\end{figure}

@@

Chick~\citep{Chick_2015} showed that by significantly relaxing the parent mass tolerance in bottom-up searches, an incredible variety of unforeseen peptides were being fragmented and could be identified.
In particular, PTM-containing peptides could be identified from those MS2 fragments that did not possess the modification and the type of PTM could be inferred from differences in precursor and parent mass.
A major limitation of the Chick work was the extremely long search times because of the vast number of theoretical candidate peptides that were evaluated for each and every MS2 spectrum.
Recently, Kong\citep{Kong_2017} reported a new index-based search strategy (MSFragger) that radically improved search speeds for wide parent mass tolerance searches.
Search times for +/- 500 Da parent tolerances went from six years to six seconds.
The output of both search strategies is a list of PSMs, where the best matching unmodified theoretical peptide was found for each MS2 fragmentation spectrum and the mass difference between the theoretical and experimental precursor masses was determined.
The mass differences can be clustered into histogram peaks\citep{Rodriguez_2014}, but interpretation of the mass difference was left to the user.
PTMs corresponding to the reported mass difference can frequently be inferred; such as when the mass difference reported matches a known PTM mass (e.g. 79.966331 Da for phosphorylation).
However, mass differences can arise from not only one modification, but frequently from two or more modifications, some of which increase the mass and the peptide and some of which decrease it. In addition, small missed cleavages are sometimes reported mistakenly as mass differences. Also, some mass differences, such as those matching to amino acid masses, are easily interpreted but suffer from high local FDR and should only be used with caution. Finally, seemingly uninterpretable mass differences can be reported when the best matching peptide has an actual precursor mass different from the precursor mass selected by the mass spectrometer. The range of masses selected for fragmentation in a tandem mass spectrometer is commonly +/- 2 Da. Therefore, multiple peptides can be introduced to the cell and co-fragmented. This often results in peptide identification of a less abundant peptide with mass slightly larger or smaller than the selected precursor.
The unconstrained utilization of mass difference and acceptance of the theoretical best matching sequence as true can result in a high number of false positive assignments.

@@

%\begin{figure}	
%	\centering
%	\begin{subfigure}[t]{3.1in}
%		\centering
%		\includegraphics[scale=0.5]{figPhosphorylationPre.png}
	%	\label{fig:1a}		
%	\end{subfigure}
%	\quad
%	\begin{subfigure}[t]{3.1in}
%		\centering
%		\includegraphics[scale=0.5]{figPhosphorylationCalib.png}
%		\label{fig:1b}
%	\end{subfigure}
%	\caption{Histograms of mass differences observed around 80 Daltons when using calibrated and uncalibrated %spectra}\label{fig:phosphoHistograms}
%\end{figure}

@@

Since the protein database is annotated with PTMs from the UniProt PTM list, it is natural to use it in the augmentation step as well.
It has a few significant drawbacks: Some mass values do not correspond to actual mass shifts, lability of modifications is not taken into account, and many PTMs are missing, including adducts and glycosylations.
Notch searches allow using expanded modification lists that include chemical modifications such as adducts and large mass PTMs such as glycosylations.
Even though adducts arising from experimental artifacts are not ultimately interesting to biologists, not including them as an option decreases the total number of identified proteins and PTMs.
A curated set of notches and the corresponding modifications is given in Appendix A.
The overall identification rate increased from 164,697 to 179,345 peptide-spectral matches (PSMs), which in turn increased the number of modified peptides by an additional 20\%.
We identified hundreds of glycosylated peptides in these unenriched samples, with many of these modifications exceeding 1000 Da.
